Better Corpus Visualization through Topic Ranking

### 2010-10-15 masters_thesis/entropy_nips_workshop/paper.tex
I reworked the entropy paper, and am close to finishing a draft

\documentclass{article}
\usepackage{nips10submit_e,times}
\usepackage{graphicx}

\title{Improving the Visualization of Topic Models}
\author{
Matthew J. Gardner \\
Department of Computer Science \\
Brigham Young University \\
Provo, UT 84604 \\
\texttt{mjg82@byu.edu} \\
\And
Other authers\ldots \\
Affiliation \\
Address \\
\texttt{email} \\
}

@@

\section{Introdcution}

Topic modeling has shown to be an effective means of extracting semantic
meaning from a collection of
documents~\cite{blei-2003-latent-dirichlet-allocation}.  Topic models have been
used as an inputs to other kinds of algorithms, such as classification and
sentiment tasks (cite some papers here), but the usefulness of topic models as
visualization tools in and of themselves has not been well established, and is
even in question (cite more papers).  Some recent work has been done in an
attempt to make corpus browsers using the results of LDA, and to make the
results of LDA more easily interpretable, but there is much work yet to be done
(still more citations).

Common approaches to using topic models to facilitate corpus browsing and
visualization are limited mainly to showing the similarity between documents in
the corpus in various ways.  Such methods are useful in their own right when
one is individually scanning documents, but they are insufficient to answer
broader questions about the corpus, like, ``What topics show up most in these
documents?'' and, ``What does a particular author in the document talk about
most frequently that is unique to that author?''  Those questions turn one away
from analyzing individual documents to trying to get information out of the
topics themselves.

@@

In order for topic models to be useful for corpus browsing, some sense has to
be made out of the topics.  One recently proposed method is to assign to each
topic a coherence value, so that the person can ignore topics that are likely
to be meaningless~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.
While topic coherence is able to eliminate meaningless topics from
consideration, it is not able to address many broader questions that a person
browsing the corpus might have.

In this paper we present the idea of using the distribution of documents for a
given topic, $p(d|z)$, in order to answer questions about a corpus that
previous visualization techniques are not able to answer.  The distribution of
topics in a document, $p(z|d)$, has been widely used in all applications of
topic modeling, from dimensionality reduction of document representations to
showing similar documents in a corpus browser.  This is understandable, given
that estimating the distribution $p(z|d)$ is one of the primary purposes of
topic modeling.  However, $p(d|z)$ also contains interesting information that
can be very useful for visualizing and understanding the output of the topic
model.

The rest of this paper is outlined as follows.  We first describe a corpus of
campaign speeches from the 2008 presidential primaries and general election.
Then we present questions that a person browsing through these campaign
speeches may want to answer, and how one can use $p(d|z)$ to answer those
questions where other methods fail.  Finally, we conclude.

\section{Campaign Speeches}

\section{What topics were most prevelant during the 2008 campaign cycle?}

\section{What topics were most universal across candidates?}

\section{What did Barack Obama talk about most frequently?}

\section{Conclusion}

>>>>

\documentclass{article}
\usepackage{nips10submit_e,times}
\usepackage{graphicx}
\usepackage{amsmath}

% Better references, I think
\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\figref}[1]{Figure~\ref{fig:#1}}
\newcommand{\algref}[1]{Algorithm~\ref{alg:#1}}

\title{Better Corpus Visualization through Topic Ranking}
\author{
Matthew J. Gardner \\
Department of Computer Science \\
Brigham Young University \\
Provo, UT 84604 \\
\texttt{mjg82@byu.edu} \\
\And
Other authers\ldots \\
Affiliation \\
Address \\
\texttt{email} \\
}

@@

\section{Introdcution}
\label{sec:introduction}

Topic modeling has been shown to be an effective means of extracting semantic
meaning from a collection of
documents~\cite{blei-2003-latent-dirichlet-allocation}.  Topic models have been
used as an inputs to other kinds of algorithms, such as classification and
sentiment tasks (cite some papers here), but the usefulness of topic models as
visualization tools in and of themselves has not been well established, and is
even in question (cite more papers).  Some recent work has been done in an
attempt to make corpus browsers using the results of LDA, and to make the
results of LDA more easily interpretable, but there is much work yet to be done
(still more citations).

Common approaches to using topic models to facilitate corpus browsing and
visualization are limited mainly to showing the similarity between documents in
the corpus in various ways.  Such methods are useful in their own right when
one is individually scanning documents, but they are insufficient to answer
broader questions about the corpus, like, ``What topics show up most in these
documents?'' and, ``What does a particular author in the corpus talk about most
frequently that is unique to that author?''  Those questions turn one away from
analyzing individual documents to trying to get information out of the topics
themselves.

@@

In order for topic models to be useful for corpus browsing, some sense has to
be made out of the topics.  One recently proposed method is to assign to each
topic a coherence value, so that the person can ignore topics that are likely
to be meaningless~\cite{newman-2010-automatic-evaluation-of-topic-coherence}.
While topic coherence is able to eliminate meaningless topics from
consideration, it is not able to address many broader questions that a person
browsing the corpus may have.

In this paper we present the novel task of ranking topics by relevance to
specific user questions.  If one is to successfully visualize a corpus in a way
that answers broad questions, and if topics, not documents, are the natural
answer to those questions, then one must properly rank topics such that the
most prominent topics contain the answer to the user's questions.  We recognize
that not all questions about a corpus fit into this visualization model, but
there are many that do, such as those listed above.

The rest of this paper is outlined as follows.  We first give further
motivation and a more detailed description of the task of ranking topics by
relevance in \secref{ranking}.  \secref{docdist} describes a method for
determining the relevance of a topic to various possible questions a person
could have when looking at a corpus, and \secref{examples} shows examples of
how this technique provides a proper ranking.  In \secref{challenges} we
discuss the challenges of evaluating this visualization method.  Finally, in
\secref{conclusion} we conclude.

\section{The topic ranking task}
\label{sec:ranking}

We assert that for many tasks, the proper visualization of a topic model is a
ranking of topics, not simply a window onto individual documents.  When one
wants to see what topics are prevelant in a corpus, the answer is not found by
scanning individual documents, but by looking at the topics produced by a topic
model.  Likewise, when one wants to see trends of topics over time, one does
not look through the documents, but at the topics, computing some statistics of
those topics for each year for which there are documents.  Thus a proper
visualization of the topic model for these tasks is to show most prominently
those topics that are most relevant to the task.

We can formalize the topic ranking problem simply as follows.  Given a set of
topics from a topic model of a corpus, and a user question about the corpus,
rank the topics by relevance to the user's question.  This is very similar to
basic information retrieval tasks, but also significantly different, in that we
are not looking for a ranking of documents as the result of a query, but a
ranking of topics which answer broader questions about the corpus as a whole.

Clearly the topic ranking task is fundamental to the proper visualization of
topic models, but as far as we are aware, this is the first attempt to even
consider the problem.  We propose a very simple way to answer some user
questions through the use of the document distribution of each topic.  This
technique is by no means a general solution to the problem, but for some
specific tasks it provides rankings that gives users the topics that answer
their questions.

\section{The document distribution for a topic}
\label{sec:docdist}

The distribution of topics in a document, $p(z|d)$, has been widely used in all
applications of topic modeling, from dimensionality reduction of document
representations to showing similar documents in a corpus browser.  This is
understandable, given that estimating the distribution $p(z|d)$ is one of the
primary purposes of topic modeling.  However, $p(d|z)$, the document
distribution for a given topic, also contains interesting information that can
be very useful for visualizing and understanding the output of the topic model.
We use various aspects of this distribution as approximations of relevance to
certain user questions.

The document distribution for a topic can be thought of as follows: suppose we
have a token labeled with a given topic; what is the probability distribution
over documents in the corpus that this token could have come from?  We
approximate this distribution with a sample from a Gibbs sampler of LDA.  The
distribution is thus defined as
\begin{align}
  \label{eq:entropy}
  p(d|z) = \frac{C(d,z)}{\sum_{d \in \mathcal{D}} C(d,z)}
\end{align}
where $d$ is a particular document, $z$ is a particular topic, $C(d,z)$ is the
number of tokens in document $d$ labeled with topic $z$, and $\mathcal{D}$ is
the corpus.

Our use of the document distribution for a topic focuses primarily on
calculating the entropy of the distribution and of related distributions where
documents are grouped by category.  For instance, one can group the documents
by author and compute the distribution of the topic over authors as
\begin{align}
  p(a|z) = \frac{\sum_{d \in \mathcal{A}} C(d,z)}{\sum_{d \in \mathcal{D}}
  C(d,z)}
\end{align}
where $a$ is an author and $\mathcal{A}$ is the set of documents written by
that author.

Also, one can compute the distribution over any subset of the documents, where
$p(d|z)$ is defined as in \eqref{eq:entropy}, but instead of the entire corpus,
$\mathcal{D}$ represents some specific subset.  When using subsets of the
documents, we only consider topics which have at least one token in the given
subset, so as to avoid division by zero.

\section{Using $p(d|z)$ to rank topics}
\label{sec:examples}

We present here a few examples of how the document distribution for a topic can
be used to rank topics according to their relevance for certain user questions.
In our examples we use a corpus of 480 campaign speeches from the 2008
presidential primary and general election.  We inferred 150 topics in the data
from a Gibbs sampling of the LDA topic model.

Questions here:

What topics are most prevelant in the corpus?

What topics did John McCain talk about the most?

What topics did Barack Obama most consistently throughout his campaign?

What topics are most consistent across candidates?

What topics were constant throughout the campaign?

What topics were transitory, only spoken of rarely?

\section{Challenges of evaluation}
\label{sec:challenges}

The obvious way to evaluate a topic ranking is to use the evaluations methods
commonly found in information retrieval tasks.  There are well established
techniques to evaluate a ranking given a true ranking.  However, there are
several obstacles to using such a scheme to evaluate performance on this task.

The first and glaring obstacle is that there is no set of data that has been
labeled with true rankings for these or any questions.  One cannot compare the
output of an algorithm to ground truth when there is no ground truth.  It is
not even clear how such data would be obtained; perhaps one could take a topic
model and a set of questions and rank each topic for relevance to each
question, but such a task would be incredibly laborious and error-prone.  The
labeler would first have to understand what exactly each topic is capturing, in
itself a difficult task, then go through the entire data collection to answer
the question herself and determine how well the topic matches.  Clearly such a
process for creating a set of evaluation data is not feasible.

But what if the topic model fails to capture topics that truly answer the
question?  
Separate topics can be overlapping in words, where a human would consider them
the same topic.

The topic model itself could be high or low quality, and that affects the
results.

\section{Conclusion}
\label{sec:conclusion}
